---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(dplyr)
library(bitfield)
library(CoordinateCleaner)
library(stringr)
library(magrittr)
library(knitr)
```

# bitfield <a href='https://github.com/ehrmanns/bitfield/'><img src='man/figures/logo.svg' align="right" height="200" /></a>

<!-- badges: start -->
<!-- [![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/)](https://cran.r-project.org/package=) -->
<!-- [![DOI](https://zenodo.org/badge/DOI/)](https://doi.org/) -->

[![R-CMD-check](https://github.com/ehrmanns/bitfield/workflows/R-CMD-check/badge.svg)](https://github.com/ehrmanns/bitfield/actions)
[![codecov](https://codecov.io/gh/ehrmanns/bitfield/branch/master/graph/badge.svg?token=hjppymcGr3)](https://codecov.io/gh/ehrmanns/bitfield)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

<!-- [![](http://cranlogs.r-pkg.org/badges/grand-total/)](https://cran.r-project.org/package=) -->
<!-- badges: end -->

## Overview

This package is designed to capture the computational footprint of any model workflow or output. It achieves this by encoding computational decisions into sequences of bits (i.e., [bitfields](https://en.wikipedia.org/wiki/Bit_field)) that are transformed to integer value. This allows storing a range of information into a single column of a table or a raster layer, which can be useful when documenting 

  - the metadata of any dataset by collecting information throughout the dataset creation process, 
  - a provenance graph that documents how a gridded modelled data product was built, 
  - intermediate data that accrue along a workflow, or 
  - a set of output metrics or parameters.

Think of a bit as a switch representing off and on states. A combination of a pair of bits can store four states, and n bits can accommodate 2^n states. These states could be the outcomes of (simple) tests that document binary responses, cases or numeric values. The data produced in that way could be described as meta-analytic or meta-algorithmic data, because they can be re-used to extend an analysis pipeline or algorithm by downstream applications.

## Installation

Install the official version from CRAN:

```{r, eval=FALSE}
# install.packages("bitfield")
```

Install the latest development version from github:

```{r, eval=FALSE}
devtools::install_github("EhrmannS/bitfield")
```


## Examples

```{r example}
library(bitfield)

library(dplyr, warn.conflicts = FALSE); library(CoordinateCleaner); library(stringr)
```

Let's first load an example dataset

```{r}
tbl_bityield$x                                       # invalid (259) and improbable (0) coordinate value

tbl_bityield$y                                       # Inf and NaN value

tbl_bityield$commodity                               # NA value or mislabelled term ("honey")

tbl_bityield$yield                                   # correct range?!

tbl_bityield$year                                    # flags (*r)

# and there is a set of valid commodity terms
validComm <- c("soybean", "maize")
```

The first step is in creating what is called `registry` in `bitfield`. This registry captures all the information required to build the bitfield

```{r}
yieldReg <- bf_registry(name = "yield_QA",
                        description = "this bitfield documents quality assessment in a table of yield data.")
```

Then, individual bit flags need to be grown by specifying the respective mapping function. These functions create flags for the most common applications, such as `bf_na()` (to test for missing values), `bf_case()` (to test what case/class the observations are part of),`bf_length()` (to count the number of digits of a variable), or `bf_numeric()` to encode a numeric (floating point) variable as bit sequence.
  
```{r}
# tests for longitude availability
yieldReg <- 
  bf_na(x = tbl_bityield,                        # specify where to determine flags
        test = "x",                              # ... and which variable to test
        pos = 1,                                 # specify at which position to store the flag
        registry = yieldReg)                     # provide the registry to update

# test which case an observation is part of
yieldReg <- 
  bf_case(x = tbl_bityield, exclusive = FALSE,
          yield >= 11, yield < 11 & yield > 9, yield < 9 & commodity == "maize",
          registry = yieldReg)

# test the length (number of digits) of values
yieldReg <- 
  bf_length(x = tbl_bityield, test = "y",
            registry = yieldReg)
  
# store a simplified (e.g. rounded) numeric value
yieldReg <-
  bf_numeric(x = tbl_bityield, source = "yield", precision = "half",
             registry = yieldReg)
```

Various derived functions build on these and thus require bits according to the same rules. The resulting data structure is a record of all the things that are grown on the bitfield.

```{r}
yieldReg
```

This is, however, not yet the bitfield. The registry is merely the instruction manual, so to speak, to create the bitfield and encode it as integer, with the function `bf_encode()`.

```{r}
(intBit <- bf_encode(registry = yieldReg))
```

The bitfield can be decoded based on the registry with the function `bf_decode()` at a later point in time, where the metadata contained in the bitfield can be studied or extended in a downstream application.

```{r}
bitfield <- bf_decode(x = intBit, registry = yieldReg, sep = "-")

# -> prints legend by default, which is also available in bf_env$legend

tbl_bityield |>
  bind_cols(bitfield) |>
  kable()
```

The column `bf_binary`, in combination with the legend, can be read one step at a time. For example, considering the first bit, we see that no observation has an `NA` value and considering the second bit, we see that observations 4 and 6 have a `yield` smaller than 9 and a `commodity` value "maize".


## Bitfields for other data-types

Not only tabular data are supported, but also gridded data such as rasters (wip).

```{r}
library(terra, warn.conflicts = FALSE)

rst_bityield <- rast(system.file("ex/rst_bityield.tif", package = "bitfield"))
levels(rst_bityield$commodity) <- tibble(id = 1:3, commodity = c("soybean", "maize", "honey"))

plot(rst_bityield)
```


# To Do

- [ ] write unit tests
- [ ] include MD5 sum for a bitfield and update it each time the bitfield is grown further
- [x] ensure everything is properly PROVy https://www.w3.org/TR/prov-o/#Activity, and document it, define the Activities that are used in the provenance statements, such as `testValue`,  `substituteValue`, `encodeAsBinary`, etc
